Predicting a Trump Victory: A State-Level Analysis of the 2024 US Presidential Election Using Bayesian Modeling

### 2024-10-31 paper/paper.qmd
committing changes to paper

---
title: "Forecasting the 2024 Election: Trump Dominates the South, Harris Leads in the Midwest"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election using state-level polling data. By aggregating high-quality polls, we predict the likely winner between Donald Trump and Kamala Harris in each state. A logistic regression model estimates the probability of Trump winning, with polling percentages as predictor variables. The model’s results highlight key battleground states and offer insights into voter dynamics, while also reflecting on the limitations of polling data in forecasting political outcomes."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(readr)
library(dplyr)
library(ggplot2)
library(usmap)
```

@@

@@

# Load cleaned data
clean_poll_data <- read_csv("/Users/jamielee/2024_US_Elections/data/02-analysis_data/analysis_data.csv")

@@

average_table <- state_average |>
  select(state, Trump_pct, Harris_pct, winner) |>
  mutate(predicted_winner = ifelse(winner == 1, "Trump", "Harris"))

@@

ggplot(stateaverage_long, aes(x=reorder(state, -support_pct), y= support_pct, fill = candidate)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black") +
  labs(x = "State", y = "Average Percentage") +
  scale_fill_manual(name = "Candidate", values = c("Trump_pct" = "red", "Harris_pct" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
The comparison shows significant variability in support for each candidate. In some states like **Montana** and **Missouri**, Trump’s support is notably higher, with a clear margin over Harris. In contrast, **Colorado** stands out as having nearly equal polling support for both candidates, suggesting a closely contested race in that state. Swing states such as **Florida**, **North Carolina**, and **Pennsylvania** exhibit relatively close polling percentages, indicating that the election outcomes in these states could be pivotal and difficult to predict. The plot also includes states like **Virginia** and **Texas**, where the support levels are more balanced, though Trump seems to hold a slight lead. The state labeled as **NA** at the end may indicate missing or incomplete data, emphasizing the need for comprehensive polling coverage in all states for more reliable predictions.

@@

The results of our predictive model suggest that Donald Trump is likely to win key states such as Florida, Missouri, and Montana, while Kamala Harris is expected to secure victories in states like Michigan, Virginia, and Pennsylvania. Close contests are predicted in battleground states like Nevada and North Carolina, where Trump holds a slight edge.

@@

Additionally, during our data cleaning process, we added constraints of high numeric grades as well as high transparency scores which led to fewer data points for us to build the model on. For example, for the state of Colorado, when we set the transparency score to higher than 5, and the numeric grade to larger and equal to 3, there are no data points for Kamala Harris. This is represented by the empty row in [@fig-averagetable].

>>>>

---
title: "Predicting a Trump Victory: A State-Level Analysis of the 2024 US Presidential Election Using Bayesian Modeling"
subtitle: "State-by-State Insights on Voter Dynamics Favoring Trump Over Harris in Key Battlgrounds"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election using state-level polling data. By aggregating high-quality polls, we predict the likely winner between Donald Trump and Kamala Harris in each state. A logistic regression model estimates the probability of Trump winning, with polling percentages as predictor variables. The model’s results highlight key battleground states and offer insights into voter dynamics, while also reflecting on the limitations of polling data in forecasting political outcomes. Our findings suggest that Trump holds an adantage in critical swing states, positinoning him as the likely winner of the election."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(usmap)
library(arrow)
```

@@

The results of our model predict a victory for Donald Trump in the 2024 U.S. Presidential Election. Key battleground states, such as Florida, Ohio, and North Carolina, lean toward Trump, emphasizing his advantage in critical regions. Our analysis points to Trump as the probable winner, with swing states playing a decisive role in determining the final outcome.

@@

# Load cleaned data
clean_poll_data <- read_parquet("/Users/jamielee/2024_US_Elections/data/02-analysis_data/analysis_data.parquet")

@@

average_table <- state_average |>
  select(
    State = state, 
    Trump = Trump_pct, 
    Harris = Harris_pct, 
    Winner = winner
    ) |>
  mutate(Winner = ifelse(Winner == 1, "Trump", "Harris"))

@@

ggplot(stateaverage_long, aes(x = reorder(state, -support_pct), y = support_pct, fill = candidate)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), color = "black") +
  labs(x = "State", y = "Average Percentage") +
  scale_fill_manual(
    name = "Candidate", 
    values = c("Trump_pct" = "red", "Harris_pct" = "blue"),
    labels = c("Trump_pct" = "Trump", "Harris_pct" = "Harris")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
The comparison shows significant variability in support for each candidate. In some states like **Montana** and **Missouri**, Trump’s support is notably higher, with a clear margin over Harris. In contrast, **Colorado** stands out as having nearly equal polling support for both candidates, suggesting a closely contested race in that state. Swing states such as **Florida**, **North Carolina**, and **Pennsylvania** exhibit relatively close polling percentages, indicating that the election outcomes in these states could be pivotal and difficult to predict. The plot also includes states like **Virginia** and **Texas**, where the support levels are more balanced, though Trump seems to hold a slight lead. The state labeled as **NA** at the end may indicate missing or incomplete data, emphasizing the need for comprehensive polling coverage in all states for more reliable predictions.

@@

The predictive model forecasts Donald Trump as the likely winner of the 2024 U.S. Presidential Election. By analyzing state-level polling data through a Bayesian logistic regression model, the study estimates the probability of Trump or Kamala Harris winning in each state. The model identifies a significant geographic split in voter support: Trump is expected to dominate in southern and midwestern states, where his polling percentages consistently surpass those of Harris. Key states projected to favor Trump include Florida, Georgia, and Ohio, which are pivotal in securing an Electoral College majority. Harris, on the other hand, holds stronger polling percentages in East Coast states and parts of the Midwest, yet these leads appear insufficient to overcome Trump’s advantage across the broader landscape of competitive states.

In particular, battleground states show varied support but lean overall toward Trump, reflecting his edge in areas like Arizona, Nevada, and North Carolina. By aggregating high-quality polls filtered for transparency and reliability, the model underscores Trump’s stable base in regions that carry substantial electoral weight. The predictions suggest that while Harris may perform well in several Democratic strongholds, Trump’s consistent support across critical swing states positions him as the probable victor in the upcoming election.

@@

During our data cleaning process, we added constraints of high numeric grades and high transparency scores, leading to fewer data points for us to build the model on. A numeric grade is a numeric rating given to the pollster indicating their quality or reliability, with a highest rating of 3.0. The transparency score is a score reflecting the pollster’s transparency about their methodology where the highest score is 10. Therefore, to ensure high-quality and reliable data to build our model on, we filtered pollsters with numeric grades of 3.0 and transparency scores of 5 or higher. This is one of the reasons why there are no data points available for Kamala Harris for the state of Colorado, represented by the empty row @fig-averagetable.

Furthermore, in the full raw data set from 538 [@FiveThirtyEight], for the state of Colorado, even without any constraints for numeric grade or transparency scores, there are only 3 data points for Kamala Harris. After reviewing the datasets, we found that except for the 3 available data points, all the rest of the polls ended in June 2024, which is a month earlier than when Biden dropped out of the race before Harris entered the presidential election race. It would not be possible that there are data points available for Harris if she was not participating in the race yet. The state of Colorado is a blue state, with the Democrats winning the state in every presidential election since 2008. Since our paper focuses on mainly the battleground states to predict election outcomes, limited data from a blue state like Colorado would not be too influential on our results.
